A Meta-Analysis of Music Emotion Recognition Studies

### 2025-04-14 manuscript/manuscript-apaquarto.qmd
polished manuscript and preprint

In the tenth annual AMC task, the highest performing model reached 69.83% accuracy [@park2017representation]. In parallel, research groups began independently evaluating MER using regression algorithms. The first study to popularize this approach predicted valence (i.e., the negative---positive emotional quality) and arousal (i.e., the calm---exciting quality) in 195 Chinese pop songs [@yang2008] using audio-extracted features represented in 114 dimensions. Applying support vector regression, the study achieved 58.3% accuracy in predicting arousal and 28.1% in predicting valence. This difference in prediction accuracy between dimensions has reappeared in several subsequent studies [e.g., @bai2016dimensional; @coutinho2013psychoacoustic], with some research suggesting this challenge reflects fewer well-established predictors and more individual differences for valence than arousal [@yang2007music; @eerola2011c].

@@

We preregistered the meta-analysis plan on 21 June 2024 at OSF, <https://osf.io/c5wgd>, and the plan is also available at <https://tuomaseerola.github.io/metaMER/preregistration/preregistration.html>).

@@

The initial search yielded 553 potential studies after excluding duplicate entries. We interactively screened them for relevance in three stages, resulting in 46 studies that passed our inclusion criteria (music emotion studies using classification or regression methods to predict emotion ratings of music using symbolic or audio features, and containing sufficient detail to convert results to $r$ or $MCC$ values see [Study Data and Code Repository - Extraction Details](https://tuomaseerola.github.io/metaMER/studies/extraction_details.html) for a breakdown). After the screening stage, we defined a set of entities to extract characterising (i) music (genre, stimulus number \[N\], duration), (ii) features extracted (number, type, source, defined by [@panda2020audio]), (iii) model type (regression, neural network, SVM, etc.) and outcome measure ($R^2$, *MSE*, *MCC*), (iv) model complexity (i.e., approximate number of features used to predict ratings), and (v) type of model cross-validation. Summary of the common datasets used in the studies in available at [Study Data and Code Repository - Datasets](https://tuomaseerola.github.io/metaMER/manuscript/datasets.html).

We converted all regression results from $R^2$ values into $r$ values for valence and arousal, and classification results into Matthews correlation coefficient [*MCC*, @chicco2020advantages]. To increase consistency in our analyses, we excluded studies using incompatible features (e.g., spectrograms of audio files [@nag2022]), or dependent variables (e.g., one regression study analyzed valence and arousal together, but not separately [@chin2018]).

![Flowchart of the study inclusions/eliminations.](figure1.svg){width="580"}

@@

The search yielded studies of variable (and occasionally questionable) quality. To mitigate potentially spurious effects resulting from the inclusion of low-quality studies, we excluded studies lacking sufficient details about stimuli, analyzed features, or model architecture [Study Data and Code Repository - Comparison](https://tuomaseerola.github.io/metaMER/studies/pass3_comparison.html). Finally, we excluded studies published in journals of questionable relevance/quality, (e.g., *Mathematical Problems in Engineering* ceased publication following 17 retractions published between July and September 2024). Overall this step eliminated 12 studies, leaving us with 34 studies in total.

@@

To capture key details of each study, we added extra fields to BibTeX entries for each study. Fields included information about the genre/type of stimuli employed, along with their duration and number; the number of analyzed features; and the model type -- Neural Nets (NN), Support Vector Machines (SVM),[^1] Linear Methods (LM), Tree-based Methods (TM), KS, Add. and KNN Methods (KM) -- validation procedure and output measures. Additionally, we included study results using executable *R* code containing custom functions for meta-analysis. For complete details about our encoding procedure, see [Study Data and Code Repository - Extraction Details](https://tuomaseerola.github.io/metaMER/studies/extraction_details.html).

@@

t <- tibble(Info = c("Study N","Model N","Techniques","Techniques","Techniques","Techniques","Techniques","Feature N","Stimulus N"), 
       Regression = c("22","204","Neural Nets: 64","Support Vector Machines: 62","Linear Methods: 62","Tree-based Methods: 14","KS, Add. & KNN: 2","Min=3, Md=653, Max=14460","Min=20, Md=324, Max=2486"),
       Classification = c("12","86","21","26","19","16","4","Min=6, Md=98, Max=8904","Min=124, Md=300, Max=5192"),
       Total = c("34","290","85","88","81","30","6","NA","NA")) 
print(knitr::kable(t,bookstabs=TRUE))
#flextable(t) %>% theme_apa()

@@

| Concept | Models, obs | $r$ \[95%-CI\] | $t$ | $p$ | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| Valence All | 102,60017 | 0.583 \[0.541-0.623\] | 21.41 | .0001 | 0.094 | 97.7% |
| Valence Max | 22,14172 | 0.669 \[0.560-0.755\] | 9.58 | .0001 | 0.148 | 98.4% |
| *N Features* |  |  |  |  |  |  |
| \<30 F | 6,3140 | 0.766 \[0.488-0.903\] | \- | \- | 0.198 | 98.6% |
| (Range: 3-22; Mdn = 16.5; *M* = 15.7) |  |  |  |  |  |  |
| 30-300 F | 8,4098 | 0.580 \[0.276-0.778\] | \- | \- | 0.188 | 97.4% |
| (Range: 45-260; Mdn = 101; *M* = 136) |  |  |  |  |  |  |
| 300+ F | 8,6934 | 0.666 \[0.531-0.767\] | \- | \- | 0.062 | 98.0% |
| (Range: 388-14460; Mdn = 654; *M* = 3360) |  |  |  |  |  |  |
| *Techniques* |  |  |  |  |  |  |
| LM | 9,2457 | 0.784 \[0.652-0.870\] | \- | \- | 0.1194 | 96.3% |
| SVM | 4,5068 | 0.539 \[0.171-0.774\] | \- | \- | 0.0702 | 97.1% |
| NN | 6,3317 | 0.473 \[0.167-0.696\] | \- | \- | 0.1029 | 98.2% |
| TM | 3,3330 | 0.750 \[0.292-0.928\] | \- | \- | 0.0740 | 98.8% |

The results indicate that valence can generally be predicted with moderately accuracy, with the best model from each of the 22 studies achieving an average correlation of *r* = 0.669 (95% CI: 0.560-0.755), called "valence Max" in Table 2. However, when considering all models across these studies (n = 102), the overall prediction rate drops significantly to *r* = 0.583. We argue that this lower correlation is likely due to the inclusion of baseline models reported in these studies, which may not reflect the true success of the task for the purposes of our analysis.

@@

To gain insights into the factors contributing to the wide range of model success, we explored several ways of splitting the data. Table 2 presents two key splits: one based on the number of features used, which we hypothesized might influence model performance, and another based on the modelling techniques employed. In terms of feature sets, we categorized them into three groups: few features (\<30), a large number of features (30–300), and massive feature sets (300+). These splits produced reasonably comparable representations of regression and classification studies, though the actual ranges differed. For transparency, we report a summary of the actual feature counts in Table 2. Models (6 in total) using a relatively small number of features (\<30) performed best (*r* = 0.766, 95% CI: 0.488–0.903) compared to those utilizing larger feature sets. However, it is worth noting that the models using massive feature sets (300+, 8 studies in total) also performed reasonably well (*r* = 0.666), achieving more consistent results than the overall prediction rate (*r* = 0.669). This observation is supported by the lowest heterogeneity index for the massive feature set group ($\tau^2$ = 0.062), indicating more consistent results across studies. Studies with large number of features (30-300 features, 8 studies in total) delivered the worst results, *r* = 0.580 (95% CI: 0.276–0.778). Despite the fluctuation in the overall model accuracy between the number of features, the differences are not substantially large to pass the test of statistical significance (*Q*(2) = 2.03, *p*=.363).

@@

These comparisons of sub-groupings may also be influenced by other factors, such as whether the study utilised predictive or explanatory modelling framework. To address this, we used the type of journal in which each study was published as a proxy indicator for predictive or explanatory modelling, where we classified studies published in psychology journals as explanatory and those published in engineering journals as predictive. Of the 22 regression studies, 13 were classified into predictive frameworks, yielding an average correlation of *r* = 0.656 (95% CI: 0.505–0.769). Studies with explanatory frameworks (9 in total) showed a similar overall correlation of *r* = 0.688 (95% CI: 0.468–0.827). No significant difference in model accuracy was observed between the two types of frameworks, *Q*(1) = 0.10, *p* = 0.748. More broadly, while the sub-groupings based on modelling techniques result in an uneven distribution of studies and observations, the two main sub-groupings presented in Table 2 highlight valuable differences in model performance across the studies. <!-- numbers in the text updated up to here, CA 20/2/2025 -->

@@

| Concept | Models, obs | $r$ \[95%-CI\] | $t$ | $p$ | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| Arousal All | 102, 60017 | 0.791 \[0.770-0.810\] | 39.9 | 0.0001 | 0.069 | 96.2% |
| Arousal Max | 22, 14172 | 0.809 \[0.740-0.860\] | 13.6 | 0.0001 | 0.141 | 97.9% |
| *N Features* |  |  |  |  |  |  |
| \<30 F | 6, 3140 | 0.885 \[0.782-0.940\] |  |  | 0.0948 | 93.5% |
| (Range: 3-22; Mdn = 16.5; *M* = 15.7) |  |  |  |  |  |  |
| 30-300 F | 8, 4098 | 0.735 \[0.501-0.868\] |  |  | 0.1971 | 98.2% |
| (Range: 45-260; Mdn = 101; *M* = 136) |  |  |  |  |  |  |
| 300+ F | 8, 6934 | 0.804 \[0.716-0.867\] |  |  | 0.0612 | 97.4% |
| (Range: 388-14460; Mdn = 654; *M* = 3360) |  |  |  |  |  |  |
| *Techniques* |  |  |  |  |  |  |
| LM | 8, 1713 | 0.882 \[0.809-0.928\] |  |  | 0.0846 | 93.3% |
| SVM | 5, 5812 | 0.796 \[0.559-0.913\] |  |  | 0.1325 | 98.3% |
| NN | 6, 3317 | 0.660 \[0.395-0.823\] |  |  | 0.1209 | 98.1% |
| TM | 3, 3330 | 0.809 \[0.733-0.864\] |  |  | 0.0025 | 65.4% |

@@

Figure 3 presents a forest plot of the random-effects model of the best-performing models from all studies. Similarly to valence, the range of correlations is also wide for arousal, ranging from 0.35 to 0.95, with all studies demonstrating evidence of positive correlations. A mean estimate of 0.81 or higher is achieved by the majority (14 out of the 22 models). Due to large sample in most studies, the confidence intervals are narrow, although the exceptions ($N < 55$) are clearly visible [@beveridge2018po; @griffiths2021am; @koh2023me; @saizclar2022pr; @wang2021ac].

@@

| Concept | Models, obs | $r$ \[95%-CI\] | $t$ | $p$ | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| All Models | 86, 80544 | 0.825 \[0.790-0.854\] | 23.7 | 0.0001 | 0.208 | 99.7% |
| Best Models | 12, 15696 | 0.868 \[0.748-0.934\] | 8.13 | 0.0001 | 0.318 | 99.8% |
| *N Features* |  |  |  |  |  |  |
| \<30 F | 4, 6179 | 0.929 \[0.816-0.974\] |  |  | 0.098 | 96.2% |
| (Range: 6-11; Mdn = 9; *M* = 8.75) |  |  |  |  |  |  |
| 30-300 F | 5, 8193 | 0.846 \[0.361-0.971\] |  |  | 0.482 | 99.9% |
| (Range: 98-231; Mdn = 122; *M* = 139) |  |  |  |  |  |  |
| 300+ F | 3, 1324 | 0.775 \[-0.271-0.982\] |  |  | 0.272 | 97.7% |
| (Range: 397-8904; Mdn = 1702; *M* = 3668) |  |  |  |  |  |  |
| *Techniques* |  |  |  |  |  |  |
| LM | 2, 735 | 0.728 \[-0.996-0.999\] |  |  | 0.194 | 98.0% |
| SVM | 3, 6556 | 0.870 \[-0.727-0.999\] |  |  | 0.823 | 99.9% |
| NN | 3, 2313 | 0.931 \[0.6523-0.988\] |  |  | 0.125 | 98.8% |
| TM | 4, 6092 | 0.853 \[0.446-0.968\] |  |  | 0.243 | 99.1% |

@@

Accounting for outliers among the classifications results removes 6 studies, affecting both performance ($MCC$ = 0.894, 95% CI: 0.828-0.936) and heterogeneity scores ($\tau^2$ = 0.057, $I^2$ = 97.7%). However, we observed no significant issues in analysis for asymmetry ($\beta$ = -19.769, 95% CI: -39.29-0.24). After aggregating across dimensions, we found that some studies with the best results only involved arousal classification [@zhang2016br; @zhang2017fe]. To assess their impact on interpretations, we evaluated how their exclusion affected average classification accuracy. These analyses revealed that accuracy ($MCC$ = 0.8758, 95%CI: 0.7235-0.9468) did not change significantly, nor did the rank order of model classes in terms of performance. Consequently, we report on all 12 in subsequent analyses.

@@

Analyzing subgroups revealed that the number of features (as classified into under 30, between 30 and 300, and above 300 features; see Table 3 for actual ranges) does not significantly impact results (*Q*(2) = 3.91, *p* = 0.1419), despite $MCC$s differing on average (\< 30 F: $MCC$ = 0.929; 30-300 F: $MCC$ = 0.846; \>300 F: $MCC$ = 0.775). Similarly, model classes did not differ significantly (*Q*(3) = 4.22, *p* = 0.239) although neural networks attained higher $MCCs$ (0.9307), followed by SVMs (0.870), tree-based methods (0.854), then linear methods (0.728). Finally, neither single vs. multigenre (*Q*(1) = 0.12, *p* = 0.732), nor binary vs. multi-class models (*Q*(1) = 0.03, *p* = 0.869) differed significantly. All classification studies were published in engineering journals and thus represent predictive framework in our earlier typology.

@@

To assess how the use of different model types affected performance, we prepared heatmap visualizations depicting differences in success across feature *n* categories and algorithms. We collapsed SVM and Tree-Based categories due to their low representation in the model summary. Figure 5 summarizes differences in success (a) across categories, as well as (b) the algorithms in each model class. Overall, studies using smaller feature sets tend to perform best, whereas the best model type largely depends on the nature of the prediction task. For valence and arousal, linear models perform better than other model types, whereas for emotion classification, neural networks show the best overall performance. The overall pattern aligns with the analyses of the splits reported earlier concerning model types and feature counts, but the visualization also highlights concurrent information about the feature n, model types, study counts, and the average number of stimuli within each combination. For instance, studies with the lowest number of features (\<30) also tend to have the highest mean number of stimuli (M = 749.5), while the poorest-performing feature count range (30–300) corresponds to the lowest mean number of stimuli (M = 522.5). It is also reassuring to observe that studies utilizing neural networks and other model types tend to use a higher number of features and stimuli than those employing linear models, as this reflects the capabilities and internal training requirements built into these models.

![Heatmaps summarizing models. Panel a depicts overall success across tasks according to model type and feature *n* summaries. Text labels summarize the number of models belonging to each subgroup, whereas the numbers in parentheses summarize the median number of stimuli used. Panel b summarizes the algorithms pertaining to each class, colour-coded by frequency. Grey rectangles are visual aids aligning with *x*-axis groupings in panel a (white: LM; light grey: NN; dark grey: other).](model-summary-fig.png "Heatmap depicting success across tasks according to Feature N category and model class. Labels indicate the number of studies and the median number of stimuli is shown in parentheses."){width="668"}

@@

We initially identified 96 studies involving MER models, but narrowed our selection to 34 after filtering studies to ensure consistent quality, reporting standards, and a specific focus on predicting emotions using music features. From these studies, we encoded accuracy scores for 290 models, with 204 related to regression and 86 to classification. Comparing the most accurate model in each study revealed reasonably accurate prediction of valence (*r* = 0.669 \[0.560, 0.755\]), and arousal (*r* = 0.809 \[0.740, 0.860\]) in regression studies. For both affective dimensions of the circumplex model, linear methods (valence *r* = 0.784, arousal *r* = 0.882) and tree-based models (valence *r* = 0.750, arousal *r* = 0.809) outperformed support vector machines (SVMs) (valence *r* = 0.539, arousal *r* = 0.796) and neural networks (valence = 0.473, arousal = 0.660). In contrast, neural networks performed most accurately in classification studies ($MCC$ = 0.931), followed by SVMs ($MCC$ = 0.870) and tree-based models ($MCC$ = 0.853). Despite the high overall success of the research in this topic, the models exhibited several differences relating to (i) the scale and quantity of data sets employed, (ii) feature extraction methods, (iii) the number and types of features and reduction methods used, (iv) the actual modelling architecture, and (v) how model outcomes are cross-validated.

@@

The results of the meta-analysis offer several important insights. First, compared to a 2013 report by @barthet2013, the glass ceiling for valence has risen from *r* = 0.51 to 0.67.[^2] Conversely, predictive accuracy for arousal has shown no improvement (reported already in 2013 at *r* = 0.83). For classification, studies reached variable classification rates from 0.497 using *F~1~* score [@sanden2011empirical] to 79% average precision [@trohidis2008multi] in the past, whereas prediction rates have improved considerably, here $MCC$ = 0.87, albeit the differences in metrics.

[^2]: Values reported as $R^2$ in original study.

@@

Broadly speaking, the present study revealed uncomfortably large variability in overall quality control and reporting in MER studies. In many cases, the reporting was insufficient to determine the features used in the models or their sources. Additionally, a significant number of studies had to be discarded due to a lack of information about the data, model architectures, or outcome measures. We summarize these issues below as recommendations for improving (1) reporting and transparency, (2) feature definition, (3) dataset scale and content, and (4) the selection of emotion frameworks.

>>>>

In the tenth annual AMC task, the highest performing model reached 69.83% accuracy [@park2017representation]. In parallel, research groups began independently evaluating MER using regression algorithms. The first study to popularize this approach predicted valence (i.e., the negative---positive emotional quality) and arousal (i.e., the calm---exciting quality) in 195 Chinese pop songs [@yang2008] using 114 audio-extracted features. Applying support vector regression, the study achieved 58.3% accuracy in predicting arousal and 28.1% in predicting valence. This difference in prediction accuracy between dimensions has reappeared in several subsequent studies [e.g., @bai2016dimensional; @coutinho2013psychoacoustic], with some research suggesting this challenge reflects fewer well-established predictors and more individual differences for valence than arousal [@yang2007music; @eerola2011c].

@@

We preregistered the meta-analysis plan on 21 June 2024 at OSF, [https://osf.io/c5wgd](https://osf.io/c5wgd), and the plan is also available at [Study and Code Repository - Preregistration](https://tuomaseerola.github.io/metaMER/preregistration/preregistration.html).

@@

The initial search yielded 553 potential studies after excluding duplicate entries. We interactively screened them for relevance in three stages, resulting in 46 studies that passed our inclusion criteria (music emotion studies using classification or regression methods to predict emotion ratings of music using symbolic or audio features, and containing sufficient detail to convert results to $r$ or $MCC$ values (see [Study Data and Code Repository - Extraction Details](https://tuomaseerola.github.io/metaMER/studies/extraction_details.html) for a breakdown). After the screening stage, we defined a set of entities to extract characterising (i) music (genre, stimulus number \[N\], duration), (ii) features extracted (number, type, source, defined by [@panda2020audio]), (iii) model type (regression, neural network, SVM, etc.) and outcome measure ($R^2$, *MSE*, *MCC*), (iv) model complexity (i.e., approximate number of features used to predict ratings), and (v) type of model cross-validation. Summary of the common datasets used in the studies in available at [Study Data and Code Repository - Datasets](https://tuomaseerola.github.io/metaMER/manuscript/datasets.html).

We converted all regression results from $R^2$ values into $r$ values for valence and arousal, and classification results into Matthews correlation coefficient [*MCC*, @chicco2020advantages]. We excluded irrelevant emotion dimensions (e.g., resonance in @deng2015em) and re-coded analogous labels (e.g., *activation* in @saizclar2022pr). Some regression studies operationalized arousal using dimensions of energy and tension [@wang2022cr]. For these studies, we excluded tension, as it has been shown to overlap considerably with valence, and with energy after partialling out the effect of valence [@eerola2011comparison]. To increase consistency in our analyses, we excluded studies using incompatible features (e.g., spectrograms of audio files in @nag2022) or dependent variables (@chin2018 evaluates valence and arousal as a single dimension).

For classification studies, the number of emotion classes ranged from two to eight. Most studies predicted classes corresponding to the affective dimensions of the circumplex model, or discrete emotion labels mapping onto them [e.g., happy, sad, nervous, and calm and similar variants used in @agarwal2021an; @alvarez2023ri; @yeh2014po]. Despite the popularity of the circumplex, its treatment varied substantially between studies. Examples range from predicting quadrants in a multi-class problem [@panda2020no] or in a series of binary classification problems [@bhuvanakumar2023em], or dividing each quadrant into multiple sublevels [@nguyen2017an; @sorussa2020em]. Some studies predicted valence and arousal separately [@hu2022de; @zhang2016br], whereas others excluded valence (e.g., all models in @zhang2017fe; the CART model in @zhang2016br) or specific quadrants (@hizlisoy2021mu excludes the bottom-right quadrant). Only a few studies classified valence and arousal separately; for these studies we averaged prediction success across both emotion concepts.

![Flowchart of the study inclusions/eliminations.](figures/figure1.png){width="580"}

@@

The search yielded studies of variable (and occasionally questionable) quality. To mitigate potentially spurious effects resulting from the inclusion of low-quality studies, we excluded studies lacking sufficient details about stimuli, analyzed features, or model architecture (see [Study Data and Code Repository - Comparison](https://tuomaseerola.github.io/metaMER/studies/pass3_comparison.html)). Finally, we excluded studies published in journals of questionable relevance/quality, (e.g., *Mathematical Problems in Engineering* ceased publication following 17 retractions published between July and September 2024). Overall this step eliminated 12 studies, leaving us with 34 studies in total.

@@

To capture key details of each study, we added extra fields to BibTeX entries for each study. Fields included information about the genre/type of stimuli employed, along with their duration and number; the number of analyzed features; and the model type -- Neural Nets (NN), Support Vector Machines (SVM),[^1] Linear Methods (LM), Tree-based Methods (TM), Kernel Smoothing, Additive and KNN Methods (KM) -- validation procedure and output measures. Additionally, we included study results using executable *R* code containing custom functions for meta-analysis. For complete details about our encoding procedure, see [Study Data and Code Repository - Extraction Details](https://tuomaseerola.github.io/metaMER/studies/extraction_details.html).

@@

t <- tibble(Info = c("Study N","Model N","Techniques","Techniques","Techniques","Techniques","Techniques","Feature N","Stimulus N"), 
       Regression = c("22","204","Neural Nets: 64","Support Vector Machines: 62","Linear Methods: 62","Tree-based Methods: 14","KS, Add. & KNN: 2","Min=3, Md=653, Max=14460","Min=20, Md=324, Max=2486"),
       Classification = c("12","86","21","26","19","16","4","Min=6, Md=98, Max=8904","Min=124, Md=300, Max=5192"),
       Total = c("34","290","85","88","81","30","6"," "," ")) 
print(knitr::kable(t,bookstabs=TRUE))
#flextable(t) %>% theme_apa()

@@

| Concept | Models, obs | $r$ \[95%-CI\] | $t$ | $p$ | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| Valence All | 102, 60017 | 0.583 \[0.541-0.623\] | 21.41 | .0001 | 0.094 | 97.7% |
| Valence Max | 22, 14172 | 0.669 \[0.560-0.755\] | 9.58 | .0001 | 0.148 | 98.4% |
| *N Features* |  |  |  |  |  |  |
| \<30 | 6, 3140 | 0.766 \[0.488-0.903\] | \- | \- | 0.198 | 98.6% |
| 30-300 | 8, 4098 | 0.580 \[0.276-0.778\] | \- | \- | 0.188 | 97.4% |
| 300+ | 8, 6934 | 0.666 \[0.531-0.767\] | \- | \- | 0.062 | 98.0% |
| *Techniques* |  |  |  |  |  |  |
| LM | 9, 2457 | 0.784 \[0.652-0.870\] | \- | \- | 0.1194 | 96.3% |
| SVM | 4, 5068 | 0.539 \[0.171-0.774\] | \- | \- | 0.0702 | 97.1% |
| NN | 6, 3317 | 0.473 \[0.167-0.696\] | \- | \- | 0.1029 | 98.2% |
| TM | 3, 3330 | 0.750 \[0.292-0.928\] | \- | \- | 0.0740 | 98.8% |

The results indicate that valence can generally be predicted with moderately accuracy, with the best model from each of the 22 studies achieving an average correlation of *r* = 0.669 (95% CI: 0.560-0.755), called "Valence Max" in Table 2. However, when considering all models across these studies (n = 102), the overall prediction rate drops significantly to *r* = 0.583. We argue that this lower correlation is likely due to the inclusion of baseline models reported in these studies, which may not reflect the true success of the task for the purposes of our analysis.

@@

To gain insights into the factors contributing to the wide range of model success, we explored several ways of splitting the data. Table 2 presents two key splits: one based on the number of features used, which we hypothesized might influence model performance, and another based on the modelling techniques employed. In terms of feature sets, we categorized them into three groups: few features (\<30), a large number of features (30–300), and massive feature sets (300+). These splits produced reasonably comparable representations of regression and classification studies, though the actual ranges differed. Models using a relatively small number of features (\<30, 6 in total) performed best (*r* = 0.766, 95% CI: 0.488–0.903) compared to those utilizing larger feature sets. However, it is worth noting that the models using massive feature sets (300+, 8 studies in total) also performed reasonably well (*r* = 0.666), achieving more consistent results than the overall prediction rate (*r* = 0.669). This observation is supported by the lowest heterogeneity index for the massive feature set group ($\tau^2$ = 0.062), indicating more consistent results across studies. Studies with large number of features (30-300 features, 8 studies in total) delivered the worst results, *r* = 0.580 (95% CI: 0.276–0.778). Despite the fluctuation in the overall model accuracy between the number of features, the differences are not substantially large to pass the test of statistical significance (*Q*(2) = 2.03, *p*=.363).

@@

These comparisons of sub-groupings may also be influenced by other factors, such as whether the study utilised predictive or explanatory modelling framework. To address this, we used the type of journal in which each study was published as a proxy indicator for predictive or explanatory modelling, where we classified studies published in psychology journals as explanatory and those published in engineering journals as predictive[^2]. Of the 22 regression studies, 13 were classified into predictive frameworks, yielding an average correlation of *r* = 0.656 (95% CI: 0.505–0.769). Studies with explanatory frameworks (9 in total) showed a similar overall correlation of *r* = 0.688 (95% CI: 0.468–0.827). No significant difference in model accuracy was observed between the two types of frameworks, *Q*(1) = 0.10, *p* = 0.748. More broadly, while the sub-groupings based on modelling techniques result in an uneven distribution of studies and observations, the two main sub-groupings presented in Table 2 highlight valuable differences in model performance across the studies. <!-- numbers in the text updated up to here, CA 20/2/2025 -->

[^2]: ASJC journal classification by Scopus was used to determine the type of subject area.

@@

| Concept | Models, obs | $r$ \[95%-CI\] | $t$ | $p$ | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| Arousal All | 102, 60017 | 0.791 \[0.770-0.810\] | 39.9 | 0.0001 | 0.069 | 96.2% |
| Arousal Max | 22, 14172 | 0.809 \[0.740-0.860\] | 13.6 | 0.0001 | 0.141 | 97.9% |
| *N Features* |  |  |  |  |  |  |
| \<30 | 6, 3140 | 0.885 \[0.782-0.940\] |  |  | 0.0948 | 93.5% |
| 30-300 | 8, 4098 | 0.735 \[0.501-0.868\] |  |  | 0.1971 | 98.2% |
| 300+ | 8, 6934 | 0.804 \[0.716-0.867\] |  |  | 0.0612 | 97.4% |
| *Techniques* |  |  |  |  |  |  |
| LM | 8, 1713 | 0.882 \[0.809-0.928\] |  |  | 0.0846 | 93.3% |
| SVM | 5, 5812 | 0.796 \[0.559-0.913\] |  |  | 0.1325 | 98.3% |
| NN | 6, 3317 | 0.660 \[0.395-0.823\] |  |  | 0.1209 | 98.1% |
| TM | 3, 3330 | 0.809 \[0.733-0.864\] |  |  | 0.0025 | 65.4% |

@@

Figure 3 presents a forest plot of the random-effects model of the best-performing models from all studies. Similarly to valence, the range of correlations is also wide for arousal, ranging from 0.35 to 0.95, with all studies demonstrating evidence of positive correlations. A mean estimate of 0.809 or higher is achieved by the majority (14 out of the 22 models). Due to large sample in most studies, the confidence intervals are narrow, although the exceptions ($N < 55$) are clearly visible [@beveridge2018po; @griffiths2021am; @koh2023me; @saizclar2022pr; @wang2021ac].

@@

| Concept | Models, obs | $r$ \[95%-CI\] | $t$ | $p$ | $\tau^2$ | $I^2$ |
|:----------|:----------|:----------|:----------|:----------|:----------|:----------|
| All Models | 86, 80544 | 0.825 \[0.790-0.854\] | 23.7 | 0.0001 | 0.208 | 99.7% |
| Best Models | 12, 15696 | 0.868 \[0.748-0.934\] | 8.13 | 0.0001 | 0.318 | 99.8% |
| *N Features* |  |  |  |  |  |  |
| \<30 | 4, 6179 | 0.929 \[0.816-0.974\] |  |  | 0.098 | 96.2% |
| 30-300 | 5, 8193 | 0.846 \[0.361-0.971\] |  |  | 0.482 | 99.9% |
| 300+ | 3, 1324 | 0.775 \[-0.271-0.982\] |  |  | 0.272 | 97.7% |
| *Techniques* |  |  |  |  |  |  |
| LM | 2, 735 | 0.728 \[-0.996-0.999\] |  |  | 0.194 | 98.0% |
| SVM | 3, 6556 | 0.870 \[-0.727-0.999\] |  |  | 0.823 | 99.9% |
| NN | 3, 2313 | 0.931 \[0.6523-0.988\] |  |  | 0.125 | 98.8% |
| TM | 4, 6092 | 0.853 \[0.446-0.968\] |  |  | 0.243 | 99.1% |

@@

Accounting for outliers among the classifications results removes 6 studies, affecting both performance ($MCC$ = 0.894, 95% CI: 0.828-0.936) and heterogeneity scores ($\tau^2$ = 0.057, $I^2$ = 97.7%). However, we observed no significant issues in analysis for asymmetry ($\beta$ = -19.77, 95% CI: -39.29-0.24). After aggregating across dimensions, we found that some studies with the best results only involved arousal classification [@zhang2016br; @zhang2017fe]. To assess their impact on interpretations, we evaluated how their exclusion affected average classification accuracy. These analyses revealed that accuracy ($MCC$ = 0.8758, 95%CI: 0.7235-0.9468) did not change significantly, nor did the rank order of model classes in terms of performance. Consequently, we report on all 12 in subsequent analyses.

@@

Analyzing subgroups revealed that the number of features (as classified into under 30, between 30 and 300, and above 300 features; see Table 3 for actual ranges) does not significantly impact results (*Q*(2) = 3.91, *p* = 0.1419), despite $MCC$s differing on average (\< 30: $MCC$ = 0.929; 30-300: $MCC$ = 0.846; 300+: $MCC$ = 0.775). Similarly, model classes did not differ significantly (*Q*(3) = 4.22, *p* = 0.239) although neural networks attained higher $MCCs$ (0.9307), followed by SVMs (0.870), tree-based methods (0.854), then linear methods (0.728). Finally, neither single vs. multigenre (*Q*(1) = 0.12, *p* = 0.732), nor binary vs. multi-class models (*Q*(1) = 0.03, *p* = 0.869) differed significantly. All classification studies were published in engineering journals and thus represent predictive framework in our earlier typology.

@@

To assess how the use of different model types affected performance, we prepared heatmap visualizations (Figure 5)depicting differences in success across feature *n* categories and algorithms. We collapsed SVM and Tree-Based categories due to their low representation in the model summary. Figure 5 summarizes differences in success (a) across categories, as well as (b) the algorithms in each model class. Overall, studies using smaller feature sets tend to perform best, whereas the best model type largely depends on the nature of the prediction task. For valence and arousal, linear models perform better than other model types, whereas for emotion classification, neural networks show the best overall performance. The overall pattern aligns with the analyses of the splits reported earlier concerning model types and feature counts, but the visualization also highlights concurrent information about the feature n, model types, study counts, and the average number of stimuli within each combination. For instance, studies with the lowest number of features (\<30) also tend to have the highest mean number of stimuli (M = 749.5), while the poorest-performing feature count range (30–300) corresponds to the lowest mean number of stimuli (M = 522.5). It is also reassuring to observe that studies utilizing neural networks and other model types tend to use a higher number of features and stimuli than those employing linear models, as this reflects the capabilities and internal training requirements built into these models.

![Heatmaps summarizing models. Panel a depicts overall success across tasks according to model type and feature *n* summaries. Text labels summarize the number of models belonging to each subgroup, whereas the numbers in parentheses summarize the median number of stimuli used. Panel b summarizes the algorithms pertaining to each class, colour-coded by success rate and the numbers in parenthesis refer to number of studies.](model-summary-fig.png "Heatmap depicting success across tasks according to Feature N category and model class. Labels indicate the number of studies and the median number of stimuli is shown in parentheses."){width="668"}

@@

We initially identified 96 studies involving MER models, but narrowed our selection to 34 after filtering studies to ensure consistent quality, reporting standards, and a specific focus on predicting emotions using music features. From these studies, we encoded accuracy scores for 290 models, with 204 related to regression and 86 to classification. Comparing the most accurate model in each study revealed reasonably accurate prediction of valence (*r* = 0.669 \[0.560, 0.755\]), and arousal (*r* = 0.809 \[0.740, 0.860\]) in regression studies. For both affective dimensions of the circumplex model, linear methods (valence *r* = 0.784, arousal *r* = 0.882) and tree-based models (valence *r* = 0.750, arousal *r* = 0.809) outperformed support vector machines (SVMs) (valence *r* = 0.539, arousal *r* = 0.796) and neural networks (valence = 0.473, arousal = 0.660). In contrast, neural networks performed most accurately in classification studies ($MCC$ = 0.931), followed by SVMs ($MCC$ = 0.870) and tree-based models ($MCC$ = 0.853). Despite the high overall success of the research in this topic, the models exhibited several differences relating to (i) the scale and quantity of data sets employed, (ii) feature extraction methods, (iii) the number and types of features and reduction methods used, (iv) the actual modelling architecture, and (v) how model outcomes were cross-validated.

@@

The results of the meta-analysis offer several important insights. First, compared to a 2013 report by @barthet2013, the prediction ceiling for valence has risen from *r* = 0.51 to 0.67.[^3] Conversely, predictive accuracy for arousal has shown no improvement (reported already in 2013 at *r* = 0.83). For classification, studies reached variable classification rates from 0.497 using *F~1~* score [@sanden2011empirical] to 79% average precision [@trohidis2008multi] in the past, whereas prediction rates have improved considerably, here $MCC$ = 0.87, albeit the differences in metrics.

[^3]: Values reported as $R^2$ in original study.

@@

Broadly speaking, the present study revealed uncomfortably large variability in overall quality control and reporting practices in MER studies. In many cases, the reporting was insufficient to determine the features used in the models or their sources. Additionally, a significant number of studies had to be discarded due to a lack of information about the data, model architectures, or outcome measures. We summarize these issues below as recommendations for improving (1) reporting and transparency, (2) feature definition, (3) dataset scale and content, and (4) the selection of emotion frameworks.
